---
title: "Storm analysis"
author: "Timo Grossenbacher"
date: "June 17, 2015"
output: html_document
---

## Task
Your data analysis must address the following questions:

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

## Synopsis

The goal of this analysis was to identify the most severe types of weather events, in terms of effects on population health and economy. To do this, the huge NOAA storm database was used. As said, the dataset contains a huge amount of events since the 50es. In order to speed up data analysis, a random sample of 10% of the original data was drawn, and only data from 2000 onwards was selected. The results of the data analysis indicate that tornadoes, heat waves and floods together cause about half of all the weather-driven fatalities. In terms of injuries, tornadoes, heat waves and strong winds seem to be responsible for more than 75% of them.

## Preparations

```{r, cache=TRUE}
# load necessary packages
if(!require(dplyr)) {
  install.packages("dplyr", repos="http://cran.us.r-project.org")
  require(dplyr)
}
if(!require(tidyr)) {
  install.packages("tidyr", repos="http://cran.us.r-project.org")
  require(tidyr)
}
if(!require(ggplot2)) {
  install.packages("ggplot2", repos="http://cran.us.r-project.org")
  require(ggplot2)
}
if(!require(magrittr)) {
  install.packages("magrittr", repos="http://cran.us.r-project.org")
  require(magrittr)
}
```

## Data Processing 

First, the data is loaded via the slightly faster `read.csv2`.
Because the data set is way to huge to process efficiently (on my very slow machine), I take a random sample of 10% of the data which should suffice to answer the questions. I also take only the data from the year 2000 onwards, for two reasons:

 * Earlier events might have caused different kinds of fatalities & damages. Society might have adopted to these, so the patterns in the data might have shifted. I only want to look at what causes problems now and not was has caused problems 50 years ago. 
 * The data might be more complete and have higher quality from the year 2000 onwards as compared to older, historical data. 

```{r, cache=TRUE}
# Unzip the data and
# Load in the data using the slightly faster read.csv2
stormdata <- read.csv2(bzfile("repdata_data_StormData.csv.bz2"), sep = ",")

set.seed(1234)
# take a sample
stormdata_reduced <- stormdata %>% sample_frac(size = 0.1, replace = F)
# in order to rule out errors due to old, unstructed data, I will only look at events occured after 2000
# first, dates need to be saved as Date objects
stormdata_reduced$BGN_DATE_AS_DATE <- as.Date(stormdata_reduced$BGN_DATE, format = "%m/%d/%Y %H:%M:%S", tz = stormdata_reduced$TIME_ZONE)
# filter
stormdata_reduced %<>% filter(BGN_DATE_AS_DATE >= as.Date("2000-01-01 00:00:00", tz = TIME_ZONE))
```

After this preprocessing, there are still over 50'000 events which should suffice for the analysis. 

Let's first explore the data a bit. How many different event types are there? 
```{r}
sort(unique(stormdata_reduced$EVTYPE))
```
There seem to be multiple problems:

 * First, the number of different event types is massive. 
 * The event types seem to be unstructured - there are duplicates (slightly differently spelt, etc.), e.g. COLD and COLD WEATHER which are essentially the same.

Now, which variables can be used to measure effects on population health? 

```{r}
str(stormdata_reduced)
unique(stormdata_reduced$FATALITIES) # seems to be numeric, so let's convert
unique(stormdata_reduced$INJURIES) # same
stormdata_reduced %<>% 
    mutate(FATALITIES = as.integer(as.character(FATALITIES)), INJURIES = as.integer(as.character(INJURIES)))
```

Seems like FATALITIES and INJURIES might help

How are they distributed? 
 
```{r}
hist(log10(stormdata_reduced$FATALITIES), main = "Frequency of fatalities", xlab = "Logarithm to the base of 10 of fatalities")
hist(log10(stormdata_reduced$INJURIES), main = "Frequency of injuries", xlab = "Logarithm to the base of 10 of injuries")
```

It seems like, for both fatalities and injuries, the distribution is heavily skewed, i.e., there are a lot of events causing no or only a few injured/dead, and only a few events causing a lot of injuries/fatalities.

Let's only look at "major" events, i.e. such over a certain threshold of injuries
```{r}
range(stormdata_reduced$FATALITIES) 
stormdata_reduced_major <- stormdata_reduced %>%
    filter(INJURIES > 3 | FATALITIES > 0) %>% tbl_df()
# there's about 450 events left
unique(stormdata_reduced_major$EVTYPE)
# still 37 different evtypes
```

Now let's group fatalities and injuries by event type 

```{r}
stormdata_reduced_major_summary <- stormdata_reduced_major %>% 
    group_by(EVTYPE) %>% 
    summarise(sum_injuries = sum(INJURIES), sum_fatalities = sum(FATALITIES))

stormdata_reduced_major_summary %>% 
    arrange(desc(sum_injuries)) %>% 
    as.data.frame()

stormdata_reduced_major_summary %>% 
    arrange(desc(sum_fatalities)) %>% 
    as.data.frame()

```
Seems like we still need to clean the EVTYPE variable
But first, EVTYPE needs to be converted to character, makes it easier

```{r}
stormdata_reduced_major %<>%
    mutate(EVTYPE = as.character(EVTYPE))
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "HEAT",]$EVTYPE <- "EXCESSIVE_HEAT"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "EXTREME COLD/WIND CHILL",]$EVTYPE <- "EXTREME COLD"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "COLD/WIND CHILL",]$EVTYPE <- "EXTREME COLD"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "COLD WEATHER",]$EVTYPE <- "EXTREME COLD"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "RIP CURRENT",]$EVTYPE <- "RIP CURRENTS"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "HIGH WIND",]$EVTYPE <- "STRONG WIND"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "THUNDERSTORM WIND",]$EVTYPE <- "STRONG WIND"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "TSTM WIND",]$EVTYPE <- "STRONG WIND"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "WIND",]$EVTYPE <- "STRONG WIND"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "WINTER WEATHER/MIX",]$EVTYPE <- "WINTER WEATHER"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "WINTER STORM",]$EVTYPE <- "BLIZZARD"
stormdata_reduced_major[stormdata_reduced_major$EVTYPE == "FLASH FLOOD",]$EVTYPE <- "FLOOD"
```



## Results

In terms of fatalities, these are the three most deadlies types of events:
```{r}
stormdata_reduced_major_summary %>% arrange(desc(sum_fatalities))  %>% slice(1:3) %>% select(EVTYPE)
cumsum(sort(stormdata_reduced_major_summary$sum_fatalities, decreasing = T))[3]/sum(stormdata_reduced_major_summary$sum_fatalities)
# They are together responsible for almost half of the fatalities
```

Let's look at the distribution of fatalities among the most deadly weather event types. 
```{r}
# sort by fatalities
fatalities <- stormdata_reduced_major_summary %>% 
    arrange(desc(sum_fatalities)) %>% 
    as.data.frame()

stormdata_reduced_major_summary %<>%
    mutate(EVTYPE = factor(EVTYPE, levels = fatalities$EVTYPE))

ggplot(fatalities %>% slice(1:10), aes(x = EVTYPE, y = sum_fatalities)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  xlab("Type of event") + ylab("Number of fatalities caused in sample dataset") +
  ggtitle("The ten most deadly weather event types")
```

In terms of injuries, these are the three most harmful types of events:
```{r}
stormdata_reduced_major_summary %>% arrange(desc(sum_injuries))  %>% slice(1:3) %>% select(EVTYPE)
cumsum(sort(stormdata_reduced_major_summary$sum_injuries, decreasing = T))[3]/sum(stormdata_reduced_major_summary$sum_injuries)
# They are together responsible for more than 3/4 of the injuries
```

Let's look at the distribution of injured among the most harmful weather event types. 
```{r}
# sort by fatalities
injuries <- stormdata_reduced_major_summary %>% 
    arrange(desc(sum_injuries)) %>% 
    as.data.frame()

stormdata_reduced_major_summary %<>%
    mutate(EVTYPE = factor(EVTYPE, levels = injuries$EVTYPE))
ggplot(injuries %>% slice(1:10), aes(x = EVTYPE, y = sum_injuries)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  theme(axis.text.x = element_text(angle=45, vjust=0.5)) +
  xlab("Type of event") + ylab("Number of injured people caused in sample dataset") +
  ggtitle("The ten most harmful weather event types")
```

## Remarks

I unfortunately didn't have time to look at economic consequences, but it is assumed that the same or similar patterns would show up: tornadoes, heat waves, floods, storms with heavy wind. 
